---
title: "Comparison between versions"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    always_allow_html: yes
bibliography: HDX.bib
vignette: >
  %\VignetteIndexEntry{Comparison between versions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(HaDeX)
library(HaDeX2)
library(microbenchmark)
library(dplyr)
library(ggplot2)
```

This article discuss differences between the first version of HaDeX, and the current version of the package and web-server. As the second version is significantly more complex than the previous one, we ultimately decided against expanding the first version. Instead we created a new entity. But how are they different?

# Methods of visualization 

Let's start by discussing methods of visualization available in the packages and web-servers:

```{r echo = FALSE}
knitr::kable(data.frame(plot_type = c("comparison", "woods", 
                     "uptake curve", "diff uptake curve", 
                     "butterfly", "diff butterfly", 
                     "chiclet", "diff chiclet", 
                     "heatmap", "diff heatmap", 
                     "3D structure",
                     "volcano", 
                     "manhattan",
                     "uncertainty",
                     "coverage", "coverage heatmap",
                     "measurement variablity", "mass uptake curve"),
           HaDeX = c(TRUE, TRUE, 
                     TRUE, FALSE,
                     FALSE, FALSE,
                     FALSE, FALSE,
                     FALSE, FALSE,
                     FALSE, 
                     FALSE, 
                     FALSE, 
                     FALSE, 
                     TRUE, FALSE, 
                     FALSE, FALSE),
           HaDeX2 = c(TRUE, TRUE,
                      TRUE, TRUE,
                      TRUE, TRUE, 
                      TRUE, TRUE,
                      TRUE, TRUE,
                      TRUE,
                      TRUE,
                      TRUE,
                      TRUE,
                      TRUE, TRUE, 
                      TRUE, TRUE)
           ))
```

# Web-server features

Let's see how the features in web-servers differ.

```{r echo=FALSE}
knitr::kable(data.frame(option = c("tooltips", "helpers", "tabular data", "times next to each other", "export to external tools"),
           HaDeX = c(TRUE, TRUE, TRUE, FALSE, FALSE), 
           HaDeX2 = c(TRUE, TRUE, TRUE, TRUE, TRUE)))
```

In the table there are shortened names. *Tabular data* means that the tabular data is available for the plot. *Times next to each other* means the option of showing all time point data in two forms: one plot or smaller plots next to each other for time point separately. *Export to external tools* means an option to download data for tools such as HDXViewer or ChimeraX.

Moreover, in HaDeX2 more visualization options are available. 

# Package functions

Finally, let's take a look at package functions from HaDeX2 and see if there is a similar function in HaDeX (even under different name)

```{r echo=FALSE}
knitr::kable(data.frame(HaDeX2 = c("add_stat_dependency", "calculate_aggregated_diff_uptake", "calculate_aggregated_test_results", "calculate_aggregated_uptake", "calculate_auc", "calculate_back_exchange", "calculate_confidence_limit_values", "calculate_diff_uptake", "calculate_exp_masses", "calculate_exp_masses_per_replicate", "calculate_kinetics", "calculate_MHP", "calculate_p_value", "calculate_peptide_kinetics", "calculate_state_uptake", "create_aggregated_diff_uptake_dataset", "create_aggregated_uptake_dataset", "create_control_dataset", "create_diff_uptake_dataset", "create_kinetic_dataset", "create_overlap_distribution_dataset", "create_p_diff_uptake_dataset", 
"create_p_diff_uptake_dataset_with_confidence", "quality_control_dataset", "create_replicate_dataset", "create_state_comparison_dataset", "create_state_uptake_dataset", "create_uptake_dataset", "get_n_replicates", "get_peptide_sequence", "get_protein_coverage", 
"get_protein_redundancy", "get_replicate_list_sd", "get_residue_positions", "get_structure_color", 
"HaDeX_GUI", "HaDeXify", "install_GUI", "plot_aggregated_differential_uptake",  "plot_aggregated_uptake", "plot_aggregated_uptake_structure", "plot_amino_distribution", 
"plot_butterfly", "plot_chiclet", "plot_coverage", "plot_coverage_heatmap", 
"plot_differential", "plot_differential_butterfly", "plot_differential_chiclet",   
"plot_differential_uptake_curve", "plot_manhattan", "plot_overlap", "plot_overlap_distribution", "plot_peptide_charge_measurement", "plot_peptide_mass_measurement", "plot_position_frequency", "plot_quality_control", "plot_replicate_histogram", "plot_replicate_mass_uptake", "plot_state_comparison", "plot_uncertainty", "plot_uptake_curve", "plot_volcano", "prepare_hdxviewer_export", "read_hdx", "reconstruct_sequence", "show_aggregated_uptake_data", 
"show_coverage_heatmap_data", "show_diff_uptake_data", "show_diff_uptake_data_confidence", "show_overlap_data", "show_p_diff_uptake_data", "show_peptide_charge_measurement", "show_peptide_mass_measurement", "show_quality_control_data", "show_replicate_histogram_data", "show_summary_data", "show_uc_data", "show_uptake_data", "update_hdexaminer_file"),
HaDeX = c("add_stat_dependency", NA, NA,NA, NA, NA, "calculate_confidence_limit_values", NA, NA, NA, "calculate_kinetics",
NA, NA, NA, NA, NA, NA, NA, NA, NA, 

NA, NA, NA, NA, NA, NA, "calculate_state_deuteration", NA, NA, NA,
NA, NA, NA, NA, NA, "HaDeX_gui", NA, NA, NA, NA,
NA, NA, NA, NA, "plot_coverage", NA, "woods_plot", NA, NA, NA,
NA, NA, "plot_position_frequency", NA, NA, NA, NA, NA, NA, "comparison_plot",

NA, "plot_kinetics", NA, NA, "read_hdx", "reconstruct_sequence", NA, NA, NA, NA,
NA, NA, NA, NA, NA, NA, NA, NA, NA, NA)))


```
For pairs of functions, we can conduct the analysis of speed using our exemplary data.


# Calculation speeds compared

Now we establish the common functionalities for both versions of the package. Using our exemplary data set, we will check the calculation times.

Here, I use outputs produced outside of this article, as I need to compare two packages with the same name.


## Reading a file

This function is called only once, in the beginning of the analysis. Here, we use an external file to call the read_hdx function. 
 
### HaDeX

```{r message = FALSE, eval=FALSE, echo = TRUE}
dat_HaDeX <- HaDeX::read_hdx(system.file(package = "HaDeX2", "HaDeX/data/alpha.csv"))
dat_HaDeX2 <- HaDeX2::read_hdx(system.file(package = "HaDeX2", "HaDeX/data/alpha.csv"))

version_benchmark <- microbenchmark(
  list = alist(`HaDeX_1. Read input` = HaDeX::read_hdx(system.file(package = "HaDeX2", "HaDeX/data/alpha.csv")),
               `HaDeX2_1. Read input` = HaDeX2::read_hdx(system.file(package = "HaDeX2", "HaDeX/data/alpha.csv")),
               `HaDeX_2. Plot uptake curve` = {dat_HaDeX %>% 
                   HaDeX::calculate_kinetics(sequence = "GFGDLKSPAGL",      
                                             state = "Alpha_KSCN", 
                                             start = 1, end = 11, 
                                             time_in = 0, time_out = 1440) %>%      
                   HaDeX::plot_kinetics(kin_dat = .)},
               `HaDeX2_2. Plot uptake curve` = {dat_HaDeX2 %>%
                   HaDeX2::calculate_peptide_kinetics(sequence = "GFGDLKSPAGL",
                                                      state = "Alpha_KSCN",
                                                      start = 1, end = 11,
                                                      time_0 = 0, time_100 = 1440) %>%
                   HaDeX2::plot_uptake_curve(uc_dat = .)})
)
```

```{r message = FALSE, echo=FALSE,results='asis'}
data.frame(version_benchmark) %>% 
   mutate(tool = sapply(strsplit(as.character(expr), "_"), first),
         task = sapply(strsplit(as.character(expr), "_"), last)) %>% 
  ggplot(aes(x = tool, y = time)) +
  geom_boxplot() +
  facet_wrap(~task, scales = "free_y")

summary(version_benchmark) %>% 
  mutate(tool = sapply(strsplit(as.character(expr), "_"), first),
         task = sapply(strsplit(as.character(expr), "_"), last)) %>% 
  select(tool, task, median, lq, uq) %>% 
  knitr::kable()

  
```
### Results

Both packages have a same-called read_hdx function. However, the backend of HaDeX2 is way more complex as there is a specific data class used, and several additional check resulting in the speed of reading and processing a file. 

## Plotting a deuterium uptake curve for one state

### HaDeX

```{r message=FALSE}

microbenchmark(
  
dat %>% 
  HaDeX::calculate_kinetics(sequence = "GFGDLKSPAGL",      
                            state = "Alpha_KSCN", 
                            start = 1, end = 11, 
                            time_in = 0, time_out = 1440) %>%      
  HaDeX::plot_kinetics(kin_dat = .)
)

```

                                                     
                                                                                                
### HaDeX2


```{r}
microbenchmark(

  dat2 %>%
    HaDeX2::calculate_peptide_kinetics(sequence = "GFGDLKSPAGL",
                               state = "Alpha_KSCN",
                               start = 1, end = 11,
                               time_0 = 0, time_100 = 1440) %>%
    HaDeX2::plot_uptake_curve(uc_dat = .)


)
```


### Results

HaDeX2 plots uptake curve significantly faster.

## Plotting a comparison plot

HaDeX plots the comparison plot only for two states.

### HaDeX 

```{r message=FALSE}

microbenchmark(
  
 HaDeX::prepare_dataset(dat = dat, 
                        in_state_first = "Alpha_KSCN_0",      
                        chosen_state_first = "Alpha_KSCN_1", 
                        out_state_first = "Alpha_KSCN_1440",      
                        in_state_second = "ALPHA_Gamma_0", 
                        chosen_state_second = "ALPHA_Gamma_1",      
                        out_state_second = "ALPHA_Gamma_1440") %>%
   HaDeX::comparison_plot(calc_dat = ., 
                          theoretical = FALSE,      
                          relative = TRUE, 
                          state_first = "Alpha_KSCN", 
                          state_second = "ALPHA_Gamma")
)

```

### HaDeX2

```{r}
microbenchmark(

  HaDeX2::create_state_comparison_dataset(dat = dat2,
                                             states = c("Alpha_KSCN", "ALPHA_Gamma"),
                                             time_0 = 0, time_100 = 1440) %>%
  HaDeX2::plot_state_comparison(uptake_dat = .,
                                          theoretical = FALSE,
                                          fractional = TRUE,
                                          time_t = 1)

)
```


### Results

The calculations are significantly faster in HaDeX2, but the difference in plotting time is resulting by new options e.q. selecting the line size, or interactive mode for GUI.

## Plotting a Woods's plot

### HaDeX

```{r message=FALSE}

microbenchmark(
  
  HaDeX::prepare_dataset(dat = dat, 
                         in_state_first = "Alpha_KSCN_0",      
                         chosen_state_first = "Alpha_KSCN_1", 
                         out_state_first = "Alpha_KSCN_1440",      
                         in_state_second = "ALPHA_Gamma_0", 
                         chosen_state_second = "ALPHA_Gamma_1",      
                         out_state_second = "ALPHA_Gamma_1440") %>%
    HaDeX::woods_plot(calc_dat = ., 
                      theoretical = FALSE,      
                      relative = TRUE, 
                      confidence_limit = 0.98, 
                      confidence_limit_2 = 0.98)
                                                                         
)

```


### HaDeX2

```{r}

microbenchmark(


  HaDeX2::create_diff_uptake_dataset(dat2,
                                     state_1 = "Alpha_KSCN",
                                     state_2 = "ALPHA_Gamma",
                                     time_0 = 0, time_100 = 1440) %>%
    HaDeX2::plot_differential(diff_uptake_dat = .,
                              time_t = 1,
                              theoretical = FALSE,
                              fractional = TRUE,
                              show_houde_interval = TRUE,
                              confidence_level = 0.98)

)
```



### Results 

As above, the plotting takes more time as it coping with option to include hybrid statistical testing, as proposed in *Hageman TS, Weis DD. Reliable Identification of Significant Differences in Differential Hydrogen Exchange-Mass Spectrometry Measurements Using a Hybrid Significance Testing Approach. Anal Chem. 2019 Jul 2;91(13):8008-8016. doi: 10.1021/acs.analchem.9b01325. Epub 2019 Jun 11. PMID: 31099554*.


# Conclusions

The HaDeX2 package is significantly more powerful than the previous version. Moreover, it was rewritten for the comfort of use - as can be shown even by the glance of parameters supplied to the function. 