---
title: "Data visualization"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    always_allow_html: yes
bibliography: HDX.bib
vignette: >
  %\VignetteIndexEntry{Data visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include = FALSE, echo = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
```

```{r, include=FALSE}
library(HaDeX)
library(ggplot2)
library(dplyr)

states <- unique(alpha_dat[["State"]])

states_uptake_dat <- create_state_comparison_dataset(alpha_dat, time_t = 1)
woods_diff_uptake_dat <- calculate_diff_uptake(alpha_dat, states = c(states[3], states[1]))
uptake_dat <- create_state_uptake_dataset(alpha_dat, state = states[3])
diff_uptake_dat <- create_diff_uptake_dataset(alpha_dat, state_1 = states[3], state_2 = states[1])
```

Data visualization is a crutial part of the experimental data analysis. The forms of visualization should be adjusted to highlight the essential result and tailored to satisfy personal needs. 

In this article, we present methods of data visualization available in the HaDeX2 package, and consecutively in the GUI.

The analyzed protein is the eEF1Bα subunit of the human guanine-nucleotide exchange factor (GEF) complex (eEF1B), measured in [Mass Spectrometry Lab](https://mslab-ibb.pl/) in [Institute of Biochemistry and Biophysics Polish Academy of Sciences](https://ibb.edu.pl/en/) and published by [Bondarchuk et al](https://doi.org/10.1093/nar/gkac685). In the one-state classification, we will focus on pure alpha state - eEF1B$\alpha$. The comparative analysis is conducted between eEF1B$\alpha$ and eEF1B$\alpha$ in presence of eEF1B$\gamma$.  

# Comparison plot

The comparison plot presents deuterium uptake of the peptides in a given time point, with information on the length of the peptide and their position in the protein sequence. It allows comparison of the results of different biological states.

*Example* In the comparison plot below, we see the fractional deuterium uptake for all three possible states: eEF1B$\alpha$, eEF1B$\alpha$ in presence of eEF1B$\gamma$ and in presence of eEF1B$\beta$, from protein eEF1B. The values are calculated for the time point 1 min. The length of the segments represents the length of the peptide and the position in the protein sequence. The error bars indicate the uncertainty of the measurement. 

```{r}
create_state_comparison_dataset(alpha_dat, time_t = 1) %>%
plot_state_comparison(., fractional = TRUE) + 
  labs(x = "Position in sequence",
       y = "Fractional deuterium uptake [%]",
       title = "Measurement after 1 min of exchange")
```


*Pros:* 
 - peptide length and position 
 - comparison of multiple biological states
 - region with difference is easy to spot on 
 - uncertainty for each peptide 

*Plot variants:*
 - fractional/absolute values
 - fractional values calculated with respect of theoretical/experimental maximal uptake
 - plots with different time points can be plotted next to each other
 - tooltips available in GUI

# Woods plot

Woods plot presents the deuterium uptake difference between two biological states for the peptides. The results are presented with respect to the length of the peptide and its position in the protein sequence for a given time point of the measurement. The statistical test (Houde et al., the confidence level is calculated based on the measurements in chosen time point of the measurement) is applied to determine the confidence limits values at the chosen level.

*Example* On the Woods plot below, we see fractional deuterium uptake difference between two biological states eEF1B$\alpha$ and eEF1B$\alpha$ in presence of eEF1B$\gamma$ for protein eEF1B The confidence limits indicate which differences are statistically significant at levels 98%. 


```{r}
calculate_diff_uptake(alpha_dat, states = c(states[3], states[1])) %>%
plot_differential(., fractional = TRUE, show_houde_interval = TRUE) +
  labs(x = "Position in seqence",
       y = "Fractional deuterium uptake difference [%]",
       title = "Measurement after 1 min of uptake")
```


*Pros* 
 - differential analysis between two states
 - peptide length and position 
 - region with difference is easy to spot on 
 - uncertainty for each peptide 
 - hybrid statistical testing
 - statistical level
 
*Plot variables:*
 - fractional/absolute values
 - fractional values calculated with respect of theoretical/experimental maximal uptake
 - plots with different time points can be plotted next to each other
 - peptides classified as statistically unsignificant can be hidden
 - tooltips available in GUI

# Butterfly plot

Butterfly plot presents the deuterium uptake for all peptides in a given state at different time points at once. Each time point of measurement is indicated by a different color. Peptides are identified by their ID (peptides are numbered arranged by the start position). 

*Example* Below, on the butterfly plot, we see how the deuterium uptake changes in time for state eEF1B$\alpha$ for protein eEF1B. We see the different exchange speed - for some peptides, the change is stable in time, and for some peptides, there is no visible change in time.

```{r}
create_state_uptake_dataset(alpha_dat, state = states[3]) %>%
plot_butterfly(., fractional = FALSE)
```

*Pros* 
 - values for time course
 - uncertainty for each peptide 

*Cons:*
 - only one state
 - lost information of length and position of the peptide
 - may be difficult to read
 
*Plot variants:*
 - fractional/absolute values
 - fractional values calculated with respect of theoretical/experimental maximal uptake
 - different methods of showing uncertainty: bars or ribbons
 - selected time points shown 
 - tooltips available in GUI
 

# Butterfly differential plot

Butterfly differential plot shows the deuterium uptake difference between two biological states in the form of a butterfly plot. It shows the results for a peptide ID (peptides are numbered arranged by the start position). The results are shown for different time points at once (time points of measurement are indicated by the color).

*Example* Below, we see how the fractional deuterium uptake difference between states eEF1B$\alpha$ and eEF1B$\alpha$ in presence of eEF1B$\gamma$ changes over time. We see that for some peptides, the difference is smaller with time - perhaps because of the back exchange.

The measurements for 1440 min are hidden, as they are close to 0, as expected.

```{r message=FALSE}
create_diff_uptake_dataset(alpha_dat, state_1 = states[3], state_2 = states[1]) %>%
  filter(Exposure < 1440) %>%
plot_differential_butterfly(fractional = TRUE, show_houde_interval = TRUE) 
```

*Pros* 
 - differential analysis between two states
 - hybrid statistical testing
 - statistical level
 - values for time course
 - uncertainty for each peptide 

*Cons:*
 - lost information of length and position of the peptide
 - may be difficult to read
 
*Plot variants:*
 - fractional/absolute values
 - fractional values calculated with respect of theoretical/experimental maximal uptake
 - different methods of showing uncertainty: bars or ribbons
 - selected time points shown 
 - tooltips available in GUI


# Chiclet plot

Chiclet plot shows the fractional deuterium uptake in the form of a heatmap for the peptides in a given biological state. One tile indicates the peptide (identified by its ID - number arranged by the start position) in a time point of measurement. The color of the tile indicates the fractinal deuterium uptake (according to the legend below the plot). 

*Example* In the chiclet plot below, we can see the deuterium uptake values for peptides (indicated by their ID) in state eEF1B$\alpha$ during the time course of the experiment. The cross symbols indicate the uncertainty of the measurement (the bigger the cross sign, the bigger the uncertainty). 

```{r}
create_state_uptake_dataset(alpha_dat, state = states[3]) %>% 
  filter(Exposure < 1440) %>%
  plot_chiclet(show_uncertainty = TRUE, fractional = FALSE)
```
*Pros* 
 - values for time course
 - uncertainty for each peptide 

*Cons:*
 - only one state
 - lost information of length and position of the peptide
 - may be difficult to read
 - small changes in values may be difficult to spot
 
*Plot variants:*
 - fractional/absolute values
 - fractional values calculated with respect of theoretical/experimental maximal uptake
 - uncertainty can be hidden to improve the readability of the plot
 - peptides classified as statistically significant can be hidden
 - selected time points shown 
 - tooltips available in GUI
 
# Chiclet differential plot

Chiclet differential plot shows the deuterium uptake difference between two biological states in the form of a heatmap. One tile indicates the peptide (identified by its ID - number arranged by the start position) in a time point of measurement. The color of the tile indicates the deuterium uptake difference (according to the legend below the plot). 

*Example* On the chiclet differential plot below, we see the fractional deuterium uptake difference between states eEF1B$\alpha$ and eEF1B$\alpha$ in presence of eEF1B$\gamma$ for protein eEF1B. We see that some peptides are protected (red), and some are deprotected (blue). The cross symbols indicate the uncertainty of the measurement (the bigger the cross sign, the bigger the uncertainty). 

```{r}
diff_uptake_dat %>%
  filter(Exposure < 1440 & Exposure > 0.001) %>%
  plot_differential_chiclet(show_uncertainty = TRUE, fractional = TRUE)
```
*Pros* 
 - differential analysis between two states
 - values for time course
 - uncertainty for each peptide 

*Cons:*
 - lost information of length and position of the peptide
 - may be difficult to read
 - small changes in values may be difficult to spot
 
*Plot variants:*
 - fractional/absolute values
 - fractional values calculated with respect of theoretical/experimental maximal uptake
 - uncertainty can be hidden to improve the readability of the plot
 - peptides classified as statistically significant can be hidden
 - selected time points shown 
 - tooltips available in GUI
 

# Volcano plot

The volcano plot shows the deuterium uptake difference for two biological states for peptide and its p-value for double testing on statistical significance (Weis et al.). On the x-axis, there is a deuterium uptake difference with its uncertainty (combined and propagated). On the y-axis, there is a P-value calculated for each peptide in a specifc time point of a measurement as a un-paired t-test on given significance level (on mass measurement from the replicates to indicate if the measured mean is significantly different between two states, as the deuterium uptake difference between states can be rewritten as 

$$\Delta D = D_{A} - D_{B} = m_{t, A} - m_{0} - (m_{t, B} - m_{0}) = m_{t, A} - m_{t, B} $$

for states A and B. The values of deuterium uptake difference from all time points are shown on the plot.

The dotted red lines indicate confidence limits for the values. The horizontal line indicates the confidence limit based on chosen confidence level to give a threshold on a P-value. The vertical lines indicate the confidence limit from Houde test for all time points and indicate a threshold on deuterium uptake difference. The statistically significant points are in the top left and right corners of the plot.

*Example* On the volcano plot below, we see the results for deuterium uptake difference between states eEF1B$\alpha$ and eEF1B$\alpha$ in presence of eEF1B$\gamma$ for protein eEF1B in all time points. The points in the left and right upper corner are statistically significant using the hybrid testing.

```{r, warning = FALSE}

p_dat <- create_p_diff_uptake_dataset(alpha_dat)

plot_volcano(p_dat, show_confidence_limits = T)

```

*Pros* This plot allows to distinguish significant results easily and uses double testing. 

*Cons* It is difficult to recognize the significant peptides and the time point of the measurement. If the plot is reactive, tooltips help to see additional information. For the not reactive plot (e.q. in articles), the information about sequence, peptide or even time point of measurement is lost. 

*Uncertainty* The uncertainty of the deuterium uptake difference (combined and propagated) is shown as the error bars in the deuterium uptake difference dimension.

*Values* This plot shows only deuterium uptake difference [Da]. 

# Uptake curve

Uptake curves show the changes in exchange in time for a specific peptide for its state.

*Example* On the uptake curve below, we see how the exchange goes for peptide GFGDLKSPAGL in all three states for protein eEF1Ba. 

```{r, warning = FALSE, message = FALSE}

calculate_peptide_kinetics(dat = alpha_dat) %>%
plot_uptake_curve() +
  ylim(c(0, NA))
```

*Pros* The plot shows the differences in deuterium uptake between two states in the time course. The back-exchange phenomenon is visible in this plot. 

*Cons* The plot shows results for only one peptide. Adding more peptides is possible but affects the readability of the plot. 

*Uncertainty* The uncertainty (combined and propagated) is shown as a ribbon. 

*Values* The values on the plot can be presented as deuterium uptake (in Daltons) or fractional deuterium uptake (in percentage). The calculation of the deuterium uptake is done with respect to experimentally measured minimal/maximal exchange control or tabular values (MHP and Max Uptake).

# Uncertainty plot

Uncertainty plot is new visualization method, showing the uncertainty of measurement of deuterium uptake for peptides to spot the regions where the uncertainty is higher. This plot may be used as a quality control of the experiment.
The presented uncertainty is in Daltons, making the threshold 1 Da as proposed limit of acceptance.

*Example* The plot below presents uncertainty values for multiple time points for state eEF1B$\alpha$. We see that the uncertainty is relatively low for all the measurement, and none of the value is suspicious.

```{r}
alpha_dat %>%
  filter(Exposure > 0) %>%
plot_uncertainty(.)
```
*Pros:* 
 - experimental quality control
 - regions with high measurement uncertainty are easy to spot 
 - peptide length and position 
 - region with difference is easy to spot on 

*Plot variants:*
 - plot for only one time point can be plotted
 - selected time points shown
 - threshold can be hidden
 - first amino omitted
 - tooltips available in GUI

# Manhattan plot

Manhattan plot is a novel plot, presenting the P-value of statistical significance between two states. 

*Example* In this example, we present the P values of difference between two biological states: eEF1B$\alpha$ and eEF1B$\alpha$ in presence of eEF1B$\gamma$.
We can see the regions where the difference is statisticaly significant - above the significance level (detailed as option in create_p_diff_uptake_dataset, or by default 0.98).

```{r}

p_diff_dat <- create_p_diff_uptake_dataset(dat = alpha_dat, diff_uptake_dat = diff_uptake_dat,
                                           state_1 = states[3], state_2 = states[1])
plot_manhattan(p_diff_dat, show_peptide_position  = TRUE)
```


*Pros* It is easy to spot regions with statistically significant changes and to observe how they change in time. 

*Cons* 

*Plot variations* Instead of peptide position and length, peptides can be identified by their ID. Only one time point can be shown. The first amino of each peptide can be skipped. The confidence level can be hidden. The tooltips are available. 

# Measurement plots

```{r}

```


*Pros* 

*Cons*

*Values*

# High-resolution plot



```{r}

kin_dat <- create_uptake_dataset(alpha_dat, states = "Alpha_KSCN")
aggregated_dat <- create_aggregated_uptake_dataset(kin_dat)
plot_aggregated_uptake(aggregated_dat)
```


*Pros* 

*Cons*

*Values*

# High-resolution on 3D structure

*Pros* 

*Cons*

*Values*

# Coverage heatmap

This plot is particulary useful when presenting AUC or back-exchange values.

AUC:
```{r}
auc_dat <- calculate_auc(create_uptake_dataset(alpha_dat))
plot_coverage_heatmap(auc_dat, value = "auc")

```

Back-exchange:
```{r}
bex_dat <- calculate_back_exchange(alpha_dat, state = "Alpha_KSCN")
plot_coverage_heatmap(bex_dat, value = "back_exchange")

```


*Pros* 

*Cons*

*Values*

# Summary of the uptake plots

Below we compare the aspects of the plots. 

```{r}

summary_plots <- data.frame(types = c("comparison", "Woods (differential)", "butterfly", "butterfly differential", "volcano", "chiclet", "chiclet differential", "uptake curve"),
                            `time course` = c(F, F, T, T, T, T, T, T),
                            `length of the peptide` = c(T, T, F, F, F, F, F, F),
                            uncertainty = c(T, T, T, T, T, T, T, T),
                            `all peptides` = c(T, T, T, T, T, T, T, F),
                            `different states` = c(T, T, F, T, T, F, T, T))

knitr::kable(summary_plots)
```


The columns indicate:

* time course - does this plot show the results from different time points?
* length of the peptide - does the plot show the information of the length of the peptide?
* uncertainty - does this plot show the uncertainty of the measurement?
* all peptides - does this plot show the results for all of the peptides?
* different states - does this plot show the results for different states?

# New features

We believe that new ways of visualization of the HDX-MS data can be developed. If so, we will be happy to include them in HaDeX package and update this article. 

