---
title: "Data visualization"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    always_allow_html: yes
bibliography: HDX.bib
vignette: >
  %\VignetteIndexEntry{Data visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include = FALSE, echo = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
```

```{r, include=FALSE}
library(HaDeX)
library(ggplot2)
library(dplyr)

states <- unique(alpha_dat[["State"]])

states_uptake_dat <- create_state_comparison_dataset(alpha_dat, time_t = 1)
woods_diff_uptake_dat <- calculate_diff_uptake(alpha_dat, states = c(states[3], states[1]))
uptake_dat <- create_state_uptake_dataset(alpha_dat, state = states[3])
diff_uptake_dat <- create_diff_uptake_dataset(alpha_dat, state_1 = states[3], state_2 = states[1])
```

Data visualization is an essential part of the data analysis as it allows to interpret the results.

In this article, we present methods of data visualization available in the HaDeX2 package. 

The analyzed protein is the eEF1Bα subunit of the human guanine-nucleotide exchange factor (GEF) complex (eEF1B), measured in [Mass Spectrometry Lab](https://mslab-ibb.pl/) in [Institute of Biochemistry and Biophysics Polish Academy of Sciences](https://ibb.edu.pl/en/) and published by [Bondarchuk et al](https://doi.org/10.1093/nar/gkac685). In the one-state classification, we will focus on pure alpha state. The comparative analysis is conducted between pure alpha state and alpha in presence of gamma component.  

# Comparison plot

The comparison plot shows deuterium uptake of the peptides in a given time point, with information on the length of the peptide and its position in the protein sequence. It allows comparing the results of different biological states.

*Example* In the comparison plot below, we see the fractional deuterium uptake for all three possible states: eEF1B$\alpha$, eEF1B$\alpha$ in presence of eEF1B$\gamma$ and in presence of eEF1B$\beta$, from protein eEF1B. The values are calculated for the time point 1 min. The length of the segments represents the length of the peptide and the position in the protein sequence. The error bars indicate the uncertainty of the measurement. 

```{r}
create_state_comparison_dataset(alpha_dat, time_t = 1) %>%
plot_state_comparison(., fractional = TRUE) + 
  labs(x = "Position in sequence",
       y = "Fractional deuterium uptake [%]",
       title = "Measurement - 1 min")
```


*Pros* The peptide deuterium uptake of each peptide can be related to its position of the protein. The changes in a specific region of the protein are visible and how it changes between the biological state at the time of the measurement.

*Cons* The information of the changes in time is lost. Some regions may be hard to read due to the overlap of peptides.

*Uncertainty* The uncertainty of the deuterium uptake measurement is indicated by the error bars associated with each peptide. 

*Values* The values on the plot can be presented as deuterium uptake (in Daltons) or fractional deuterium uptake (in percentage). The calculation of the deuterium uptake is done with respect to experimentally measured minimal/maximal exchange control or tabular values (MHP and Max Uptake).

# Woods plot

Woods plot presents the deuterium uptake difference between two biological states for the peptides. The results are presented with respect to the length of the peptide and its position in the protein sequence for a given time point of the measurement. The statistical test (Houde et al., the confidence level is calculated based on the measurements in chosen time point of the measurement) is applied to determine the confidence limits values at the chosen level.

*Example* On the Woods plot below, we see fractional deuterium uptake difference between two biological states eEF1B$\alpha$ and eEF1B$\alpha$ in presence of eEF1B$\gamma$ for protein eEF1B The confidence limits indicate which differences are statistically significant at levels 98%. 


```{r}
calculate_diff_uptake(alpha_dat, states = c(states[3], states[1])) %>%
plot_differential(., fractional = TRUE, show_houde_interval = TRUE) +
  labs(x = "Position in seqence",
       y = "Fractional deuterium uptake difference [%]",
       title = "Measurement - 1 min")
```


*Pros* The peptide deuterium uptake difference between two biological states of each peptide can be related to its position of the protein. The changes in a specific region of the protein are visible. The confidence limits indicate if the result is statistically significant. 

*Cons* The information of the changes in time is lost. Some regions may be hard to read due to the overlap of peptides.

*Uncertainty* The uncertainty of the deuterium uptake difference (uncertainty of the measurement combined and propagated) is indicated by the error bars associated with each peptide. 

*Values* The values on the plot can be presented as deuterium uptake difference (in Daltons) or fractional deuterium uptake difference (in percentage). The calculation of the deuterium uptake difference is done with respect to experimentally measured minimal/maximal exchange control or tabular values (MHP and Max Uptake).

# Butterfly plot

Butterfly plot presents the deuterium uptake for all peptides in a given state at different time points at once. Each time point of measurement is indicated by a different color. Peptides are identified by their ID (peptides are numbered arranged by the start position). 

*Example* Below, on the butterfly plot, we see how the deuterium uptake changes in time for state eEF1B$\alpha$ for protein eEF1B. We see the different exchange speed - for some peptides, the change is stable in time, and for some peptides, there is no visible change in time.

```{r}
create_state_uptake_dataset(alpha_dat, state = states[3]) %>%
plot_butterfly(., fractional = FALSE)
```

*Pros* Shows the changes of deuterium uptake in time and allows to see how the exchange goes. 

*Cons* It shows the results only for one biological state. The information on the length of the peptide and the protein region is lost.

*Uncertainty* Uncertainty of the deuterium uptake (combined and propagated based on the chosen value) is shown as a ribbon or error bars.

*Values* The values on the plot can be presented as deuterium uptake (in Daltons) or fractional deuterium uptake (in percentage). The calculation of the deuterium uptake is done with respect to experimentally measured minimal/maximal exchange control or tabular values (MHP and Max Uptake).

# Butterfly differential plot

Butterfly differential plot shows the deuterium uptake difference between two biological states in the form of a butterfly plot. It shows the results for a peptide ID (peptides are numbered arranged by the start position). The results are shown for different time points at once (time points of measurement are indicated by the color).

*Example* Below, we see how the fractional deuterium uptake difference between states eEF1B$\alpha$ and eEF1B$\alpha$ in presence of eEF1B$\gamma$ changes over time. We see that for some peptides, the difference is smaller with time - perhaps because of the back exchange.

```{r message=FALSE}
plot_differential_butterfly(diff_uptake_dat, fractional = TRUE) 
```

*Pros* It shows the results for different time points of the measurement. 

*Cons* The readability of the plot is very small due to small and overlapping values. There is no indication which changes are statistically significant. The information of peptide length and protein region is lost.

*Uncertainty*  Uncertainty of the deuterium uptake difference (combined and propagated based on the chosen value) is shown as a ribbon or error bars.

*Values* The values on the plot can be presented as deuterium uptake difference (in Daltons) or fractional deuterium uptake difference (in percentage). The calculation of the deuterium uptake difference is done with respect to experimentally measured minimal/maximal exchange control or tabular values (MHP and Max Uptake).

# Chiclet plot

Chiclet plot shows the fractional deuterium uptake in the form of a heatmap for the peptides in a given biological state. One tile indicates the peptide (identified by its ID - number arranged by the start position) in a time point of measurement. The color of the tile indicates the fractinal deuterium uptake (according to the legend below the plot). 

*Example* In the chiclet plot below, we can see the deuterium uptake values for peptides (indicated by their ID) in state eEF1B$\alpha$ during the time course of the experiment. The cross symbols indicate the uncertainty of the measurement (the bigger the cross sign, the bigger the uncertainty). 

```{r}
uptake_dat %>%
  filter(Exposure < 1440) %>%
  plot_chiclet(show_uncertainty = T, fractional = TRUE)
```

*Pros* The plot shows how the deuterium uptake changes over time. 

*Cons* This plot shows the values only for one biological state. The values are indicated by colors, so the small changes between time points are difficult to notice. The information about peptide length and protein region is lost. 

*Uncertainty*  Uncertainty (combined and propagated) is shown as the cross symbols of each tile. The legend is placed under the plot. 

*Values* The values on the plot can be presented as deuterium uptake (in Daltons) or fractional deuterium uptake (in percentage). The calculation of the deuterium uptake is done with respect to experimentally measured minimal/maximal exchange control or tabular values (MHP and Max Uptake).

# Chiclet differential plot

Chiclet differential plot shows the deuterium uptake difference between two biological states in the form of a heatmap. One tile indicates the peptide (identified by its ID - number arranged by the start position) in a time point of measurement. The color of the tile indicates the deuterium uptake difference (according to the legend below the plot). 

*Example* On the chiclet differential plot below, we see the fractional deuterium uptake difference between states eEF1B$\alpha$ and eEF1B$\alpha$ in presence of eEF1B$\gamma$ for protein eEF1B. We see that some peptides are protected (red), and some are deprotected (blue). The cross symbols indicate the uncertainty of the measurement (the bigger the cross sign, the bigger the uncertainty). 

```{r}
diff_uptake_dat %>%
  filter(Exposure < 1440 & Exposure > 0.001) %>%
  plot_differential_chiclet(show_uncertainty = TRUE, fractional = TRUE)
```

*Pros* The plot shows the results for different time points of the measurement.

*Cons* The values are indicated by colors, so the small changes between time points are difficult to notice. There is no indication if the values are statistically significant. If we choose not to show the insignificant values, that may be mistaken for no difference. The information on peptide length and protein region is lost. 

*Uncertainty* Uncertainty (combined and propagated) is shown as the cross symbols of each tile. The legend is placed under the plot. 

*Values* The values on the plot can be presented as deuterium uptake difference (in Daltons) or fractional deuterium uptake difference (in percentage). The calculation of the deuterium uptake difference is done with respect to experimentally measured minimal/maximal exchange control or tabular values (MHP and Max Uptake).

# Volcano plot

The volcano plot shows the deuterium uptake difference for two biological states for peptide and its p-value for double testing on statistical significance (Weis et al.). On the x-axis, there is a deuterium uptake difference with its uncertainty (combined and propagated). On the y-axis, there is a P-value calculated for each peptide in a specifc time point of a measurement as a un-paired t-test on given significance level (on mass measurement from the replicates to indicate if the measured mean is significantly different between two states, as the deuterium uptake difference between states can be rewritten as 

$$\Delta D = D_{A} - D_{B} = m_{t, A} - m_{0} - (m_{t, B} - m_{0}) = m_{t, A} - m_{t, B} $$

for states A and B. The values of deuterium uptake difference from all time points are shown on the plot.

The dotted red lines indicate confidence limits for the values. The horizontal line indicates the confidence limit based on chosen confidence level to give a threshold on a P-value. The vertical lines indicate the confidence limit from Houde test for all time points and indicate a threshold on deuterium uptake difference. The statistically significant points are in the top left and right corners of the plot.

*Example* On the volcano plot below, we see the results for deuterium uptake difference between states eEF1B$\alpha$ and eEF1B$\alpha$ in presence of eEF1B$\gamma$ for protein eEF1B in all time points. The points in the left and right upper corner are statistically significant using the hybrid testing.

```{r, warning = FALSE}

p_dat <- create_p_diff_uptake_dataset(alpha_dat)

plot_volcano(p_dat, show_confidence_limits = T)

```

*Pros* This plot allows to distinguish significant results easily and uses double testing. 

*Cons* It is difficult to recognize the significant peptides and the time point of the measurement. If the plot is reactive, tooltips help to see additional information. For the not reactive plot (e.q. in articles), the information about sequence, peptide or even time point of measurement is lost. 

*Uncertainty* The uncertainty of the deuterium uptake difference (combined and propagated) is shown as the error bars in the deuterium uptake difference dimension.

*Values* This plot shows only deuterium uptake difference [Da]. 

# Uptake curve

Uptake curves show the changes in exchange in time for a specific peptide for its state.

*Example* On the uptake curve below, we see how the exchange goes for peptide GFGDLKSPAGL in all three states for protein eEF1Ba. 

```{r, warning = FALSE, message = FALSE}

calculate_peptide_kinetics(dat = alpha_dat) %>%
plot_uptake_curve() +
  ylim(c(0, NA))
```

*Pros* The plot shows the differences in deuterium uptake between two states in the time course. The back-exchange phenomenon is visible in this plot. 

*Cons* The plot shows results for only one peptide. Adding more peptides is possible but affects the readability of the plot. 

*Uncertainty* The uncertainty (combined and propagated) is shown as a ribbon. 

*Values* The values on the plot can be presented as deuterium uptake (in Daltons) or fractional deuterium uptake (in percentage). The calculation of the deuterium uptake is done with respect to experimentally measured minimal/maximal exchange control or tabular values (MHP and Max Uptake).

# Uncertainty plot

Uncertainty plot is new visualization method, showing the uncertainty of measurement of deuterium uptake for peptides to spot the regions where the uncertainty is higher. This plot may be used as a quality control of the experiment.
The presented uncertainty is in Daltons, making the threshold 1 Da as proposed limit of acceptance.

*Example* The plot below presents uncertainty values for multiple time points for state eEF1B$\alpha$. We see that the uncertainty is relatively low for all the measurement, and none of the value is suspicious.

```{r}
alpha_dat %>%
  filter(Exposure > 0) %>%
plot_uncertainty(.)
```
*Pros* We see the uncertainty values for peptides with information about their length and position in the whole protein sequecene, making it easy to spot regions with higher uncertainty. In presented form, we can compare the change in time.

*Cons* 

*Plot variations* It is possible to show values only for one selected time points. There is also option to omit first amino. The threshold of 1 Da can be hidden. The tooltips are available if needed. 

# Manhattan plot

Manhattan plot is a novel plot, presenting the P-value of statistical significance between two states. 

*Example* In this example, we present the P values of difference between two biological states: eEF1B$\alpha$ and eEF1B$\alpha$ in presence of eEF1B$\gamma$.
We can see the regions where the difference is statisticaly significant - above the significance level (detailed as option in create_p_diff_uptake_dataset, or by default 0.98).

```{r}

p_diff_dat <- create_p_diff_uptake_dataset(dat = alpha_dat, diff_uptake_dat = diff_uptake_dat,
                                           state_1 = states[3], state_2 = states[1])
plot_manhattan(p_diff_dat, show_peptide_position  = TRUE)
```


*Pros* It is easy to spot regions with statistically significant changes and to observe how they change in time. 

*Cons* 

*Plot variations* Instead of peptide position and length, peptides can be identified by their ID. Only one time point can be shown. The first amino of each peptide can be skipped. The confidence level can be hidden. The tooltips are available. 

# Measurement plots

```{r}

```


*Pros* 

*Cons*

*Values*

# High-resolution plot



```{r}

kin_dat <- create_uptake_dataset(alpha_dat, states = "Alpha_KSCN")
aggregated_dat <- create_aggregated_uptake_dataset(kin_dat)
plot_aggregated_uptake(aggregated_dat)
```


*Pros* 

*Cons*

*Values*

# High-resolution on 3D structure

*Pros* 

*Cons*

*Values*

# Coverage heatmap

This plot is particulary useful when presenting AUC or back-exchange values.

AUC:
```{r}
auc_dat <- calculate_auc(create_uptake_dataset(alpha_dat))
plot_coverage_heatmap(auc_dat, value = "auc")

```

Back-exchange:
```{r}
bex_dat <- calculate_back_exchange(alpha_dat, state = "Alpha_KSCN")
plot_coverage_heatmap(bex_dat, value = "back_exchange")

```


*Pros* 

*Cons*

*Values*

# Summary of the uptake plots

Below we compare the aspects of the plots. 

```{r}

summary_plots <- data.frame(types = c("comparison", "Woods (differential)", "butterfly", "butterfly differential", "volcano", "chiclet", "chiclet differential", "uptake curve"),
                            `time course` = c(F, F, T, T, T, T, T, T),
                            `length of the peptide` = c(T, T, F, F, F, F, F, F),
                            uncertainty = c(T, T, T, T, T, T, T, T),
                            `all peptides` = c(T, T, T, T, T, T, T, F),
                            `different states` = c(T, T, F, T, T, F, T, T))

knitr::kable(summary_plots)
```


The columns indicate:

* time course - does this plot show the results from different time points?
* length of the peptide - does the plot show the information of the length of the peptide?
* uncertainty - does this plot show the uncertainty of the measurement?
* all peptides - does this plot show the results for all of the peptides?
* different states - does this plot show the results for different states?

# New features

We believe that new ways of visualization of the HDX-MS data can be developed. If so, we will be happy to include them in HaDeX package and update this article. 

